---
title: "Anhydrobiosis Index"
author: "Matteo Vecchi"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
output: 
  workflowr::wflow_html:
    code_folding: hide 
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r results='hide', message=F, warning=F}
library(tidyverse)
```

## Modelling recovery from anhydrobiosis with an exponential distribution

Most of the studies involving the quantification of anhydrobiotic performances in tardigrades measure the proportion of moving animals in different times after the beginning of rehydration. The time to recovery of individuals after the beginning the rehydration can be described by an **exponential distribution**, that is the probability distribution of the time between events that occur continuously and independently at a constant average rate.\
The probability distribution funtion (PDF) of the exponential distribution function models the time to event, however usually this information is not available as it would require to constantly observe the individuals and record the exact moment when they start moving. It is more common instead to check the individuals at fixed time points and record how many of them are moving. The proportion if individuals that recovered their motility can be instead modeled with the cumulative distribution function (CDF) of the exponential distribution.

```{r Exponential distribution function, fig.dim = c(8, 4.5)}
x = seq(0,5,0.01)
lambda = 1
y_PDFexp = lambda*exp(-lambda*x)
y_CDFexp = 1 - exp(-lambda*x)
par(mfrow=c(1,2))
plot(x=x, y=y_PDFexp, type="l", xlab="x",ylab="P(x)", main = "Probability density function (PDF)", cex.axis=0.75, cex.main=0.9)
plot(x=x, y=y_CDFexp, type="l", xlab="x",ylab="P(X <= x)", main = "Cumulative distribution function (CDF)", cex.axis=0.75, cex.main=0.9)
```

The PDF (image above - left) describes how much time do we have to wait to observe an individual starting to move after the rehydration. The CDF (image above - right) describes instead what proportion of animals already started moving at a given time *x*. From now on we will discuss only about the CDF as it is more fit to the data that is usually produce in the tardigrade anhydrobiosis experiments.

The exponential distribution is only describe by one parameter $\lambda$ (lambda), often called the rate parameter and can take any positive value. $\frac{1}{\lambda}$ is the average time that we have to wait to observe the event of interest. The CDF of an exponential distribution has the following formula (where *x* represents time): $$ P(x \le X) = 1 - e^{-\lambda*x}$$

```{r CDF different lambdas, fig.dim = c(8, 3)}
par(mfrow=c(1,3))
x = seq(0,5,0.01)
y = 1 - exp(-0.25*x)
plot(x=x, y=y, type="l", xlab="x",ylab="P(X <= x)", main = "CDF, lambda = 0.25", cex.axis=0.75, cex.main=0.9)
y = 1 - exp(-1*x)
plot(x=x, y=y, type="l", xlab="x",ylab="P(X <= x)", main = "CDF, lambda = 1.0", cex.axis=0.75, cex.main=0.9)
y = 1 - exp(-1.75*x)
plot(x=x, y=y, type="l", xlab="x",ylab="P(X <= x)", main = "CDF, lambda = 1.75", cex.axis=0.75, cex.main=0.9)
```

With smaller $\lambda$ values, the curve grows slower (the individuals takes longer on average to recover), whereas with bigger $\lambda$ values, the curve grows faster.

The exponential distribution also assumes that at one point in time, all individuals will recover. This is not realistic as the anhydrobiotic survival can be lower than 100%. This can be easily accommodated by multiplying by the final survival. 
$$ p = P(Survival) * (1 - e^{-\lambda*x})$$
```{r CDF different p, fig.dim = c(8, 3)}
par(mfrow=c(1,3))
x = seq(0,5,0.01)
y = (1 - exp(-1*x))*1
plot(x=x, y=y, type="l", xlab="x",ylab="p", main = "CDF, lambda = , P(survival) = 1", cex.axis=0.75, cex.main=0.9, ylim=c(0,1))
y = (1 - exp(-1*x))*0.75
plot(x=x, y=y, type="l", xlab="x",ylab="p", main = "CDF, lambda = , P(survival) = 0.75", cex.axis=0.75, cex.main=0.9, ylim=c(0,1))
y = (1 - exp(-1*x))*0.5
plot(x=x, y=y, type="l", xlab="x",ylab="p", main = "CDF, lambda = , P(survival) = 0.50", cex.axis=0.75, cex.main=0.9, ylim=c(0,1))
```
By changing the two parameters ($\lambda$ and *p*) of this modified CDF of exponential distribution, we can describe accurately how does tardigrades start recovering after anhydrobiosis. 

If we measured the proportion of motile animals at 1,2,4,12,and 24 hours after the start of the rehydration (red dots), we would fit all those single points to a unique line (black line).  

```{r data fitting}
x = seq(0,25,0.1)
y = (1 - exp(-0.3*x))*0.75
plot(x=x, y=y, type="l", xlab="time(hours)",ylab="Proportion of motile animals", main = "Fitting of anhydrobiosis recovery data", 
     cex.axis=0.75, cex.main=0.9, ylim=c(0,1))
points(x=c(1,2,4,12,24), y=c(0.22,0.30,0.50,0.77,0.74), col="red", pch=19)
```

If we fit the line with Bayestan methods, we will obtain not one single line, but a posterior distribution (grey lines) of equally probable lines (of which we can calculate the average one - black line).  
```{r bayesian fitting}
x = seq(0,25,0.1)
y_true = (1 - exp(-0.3*x))*0.75
plot(x=x, y=y_true, type="n", xlab="time(hours)",ylab="Proportion of motile animals", main = "Fitting of anhydrobiosis recovery data", 
     cex.axis=0.75, cex.main=0.9, ylim=c(0,1))

for(i in 1:100){
  y = (1 - exp(-(0.3+rnorm(1,0,0.05))*x))*(0.75+rnorm(1,0,0.05))
  lines(x=x, y=y, type="l", col =alpha("grey", 0.5))
  
}
lines(x=x, y=y_true)
points(x=c(1,2,4,12,24), y=c(0.22,0.30,0.50,0.77,0.74), col="red", pch=19)
```

## Reparametrization of CDF exponential function
The $\lambda$ parameter is not easy to interpret. I t is possible however, to reparametrize the function to a new parameter that describe the time needed to reach the motility of a certain proportion of the surviving animals.  
The new parameter (that we will call *Q~pm~*), describe the time needed to reach a certain proportion (*pm*) proportion of animals to be motile:
$$ \lambda = \frac{-log(1-pm)}{Q} $$
If we set *pm* = 0.9, and then *Q~0.9~* is 12, it means that (among the surviving animals), 90% of them will recover their motility in 12 hours after rehydration. 
With *pm* set to 0.9, and assuming a survival of 100%, the different recovery speed would look like this:  

```{r different Qs}
x = seq(0,25,0.1)
plot(x=c(0,25), y=c(0,1), type="n", xlab="time(hours)",ylab="Proportion of motile animals", main = "", 
     cex.axis=0.75, cex.main=0.9, ylim=c(0,1), xaxt="n")
axis(side=1, at=c(1,4,8,12,16,20,24), labels = TRUE)

for(Q in c(1,4,8,12,16,20,24)){
  lambda = -log(1-0.9)/Q
  y = (1 - exp(-lambda*x))*1
  lines(x=x, y=y, type="l", col ="black")
}
abline(h = 0.9, col="red", lty=2)
abline(v = c(1,4,8,12,16,20,24), col="blue", lty=2)
axis(3, at = c(1,4,8,12,16,20,24), 
     labels = paste0("Q0.9 = ",c(1,4,8,12,16,20,24)), 
     lwd = 0, col.axis = "blue",cex.axis=0.75)


```

The same keeps whrn the survival is lower than 100%, the 90% quantile is calculated on the estimated survival. For example, with a final survival of 75%, the 90% quantile for *Q~0.9~* is calculated as $0.75*0.90 = 0.675$.  

```{r different QsP75}
x = seq(0,25,0.1)
plot(x=c(0,25), y=c(0,1), type="n", xlab="time(hours)",ylab="Proportion of motile animals", main = "", 
     cex.axis=0.75, cex.main=0.9, ylim=c(0,1), xaxt="n")
axis(side=1, at=c(1,4,8,12,16,20,24), labels = TRUE)

for(Q in c(1,4,8,12,16,20,24)){
  lambda = -log(1-0.9)/Q
  y = (1 - exp(-lambda*x))*0.75
  lines(x=x, y=y, type="l", col ="black")
}
abline(h=0.75, col="green")
abline(h = 0.9*0.75, col="red", lty=2)
abline(v = c(1,4,8,12,16,20,24), col="blue", lty=2)
axis(3, at = c(1,4,8,12,16,20,24), 
     labels = paste0("Q0.9 = ",c(1,4,8,12,16,20,24)), 
     lwd = 0, col.axis = "blue",cex.axis=0.75)
```

With this formulation, *Q~pm~* is independent from the overall survival probability and has an intuitive interpretation.  

Note that *Q~pm~* with different *pm* can be converted to each other with the following formula:
$$Q_{new} = \frac{log(1-pm_{new})}{log(1-pm_{old})}*Q_{old} $$
For example, following this formula $Q_{0.9}=20$ and $Q_{0.5}=6.02$ are equivalent and represent only different ways to describe the same recovery speed.  

## Priors for Q~pm~ and p in Bayesian estimation
As we are estimating the *p* and *Q~pm~* parameters with Bayesian methods, we have to specify priors for them. The priors allow us to bind their possible value to only the onses that we could observe. The *p* parameter is a probability, so it must be bound between 0 and 1, whereas *Q~pm~* is expressed in hours, so it can only take positive values.  
For *p* we can set au uniform prior between 0 and 1.
$$ p \sim Dunif(0,1)$$
For *Q~pm~*, a truncated (only positive values) Gaussian prior centered at 0 and a high standard deviation will be used. This will ensure a low informative prior in the range of biologically plausible values for *Q~pm~* (0-48), while avoiding drifting to unrealistic higher values.  
$$ Q{pm} \sim N(0,100)T(0,)$$
```{r Qpm priors, fig.dim = c(8, 4.5)}
par(mfrow=c(1,2))
plot(density(rnorm(10000,0,100), adjust=1.5), xlim=c(0,500), xlab="hours", main="N(0,100)T(0,)")
plot(density(rnorm(10000,0,100), adjust=1.5), xlim=c(0,48), xlab="hours", main="N(0,100)T(0,) - Zoom")
```
Note that in the 0-48 hours range the prior is almost flat.  

## From *p* and *Q~pm~* to Anhydrobiosis Index

Summarizing both *p* and *Q~pm~* in one single index would provide a simple statistics to use for statistical analysis. The index would ideally be bound between 0 and 1, to be easily comparable across studies and interpretable.  
A good solution is to use the area under the estimated motility recovery curve, divided by the maximum value it could take if all the individuals would survive and start moving immediately after rehydration. 
The time interval used for the calculation of the Anhydrobiosis Index must be kept constant (as it will change if, for example, a maximum of 48h is used instead then 24).

```{r, fig.dim = c(8, 4.5)}
par(mfrow=c(1,2))
plot(x= c(0,24), y=c(0,1), type="n", xlab ="time (hours)", ylab="Proportion of motile individuals",
     main = "Maximum curve area",cex.axis=0.75, cex.main=0.9, ylim=c(0,1))
polygon(x = c(0,0,24,24),                 # X-Coordinates of polygon
        y = c(0,1,1,0),          # Y-Coordinates of polygon
        col = "light blue") 

x = seq(0,24,0.1)
y = (1 - exp(-0.3*x))*0.75
plot(x=c(0,24), y=c(0,1), type="n", xlab="time(hours)",ylab="Proportion of motile animals", main = "Estimated curve area", 
     cex.axis=0.75, cex.main=0.9, ylim=c(0,1))
polygon(x=c(x,24), y=c(y,0), col =alpha("red", alpha=0.5))
```

$$ AH_{24} = \frac{\color{Red} {\text{Estimated area under the curve (0-24h)}}}{\color{Blue}{\text{Maximum area under the curve (0-24h)}}} $$

With this formulation, *AH~24~* can be described as the survival probability *p*, penalized by the speed of recovery *Q~pm~* (the lower the speed, the more the penalization). By increasing the time-frame on which the index is calculated, it will approximate the *p* more closely.  

```{r AH approximates p, fig.dim = c(8, 4.5)}
par(mfrow=c(1,2))
#Ah24
plot(x= c(0,24), y=c(0,1), type="n", xlab ="time (hours)", ylab="Proportion of motile individuals",
     main = expression("p = 0.75, Q"[0.9]*" = 20, AH"[24]*" = 0.49"),cex.axis=0.75, cex.main=0.9, ylim=c(0,1))
polygon(x = c(0,0,24,24),                 # X-Coordinates of polygon
        y = c(0,1,1,0),          # Y-Coordinates of polygon
        col = "light blue") 

lambda = -log(1-0.9)/20
x = seq(0,24,0.1)
y = (1 - exp(-lambda*x))*0.75
polygon(x=c(x,24), y=c(y,0), col =alpha("red", alpha=0.5))

# AH48
plot(x= c(0,48), y=c(0,1), type="n", xlab ="time (hours)", ylab="Proportion of motile individuals",
     main = expression("p = 0.75, Q"[0.9]*" = 20, AH"[48]*" = 0.61"),cex.axis=0.75, cex.main=0.9, ylim=c(0,1))
polygon(x = c(0,0,48,48),                 # X-Coordinates of polygon
        y = c(0,1,1,0),          # Y-Coordinates of polygon
        col = "light blue") 

lambda = -log(1-0.9)/20
x = seq(0,48,0.1)
y = (1 - exp(-lambda*x))*0.75
polygon(x=c(x,48), y=c(y,0), col =alpha("red", alpha=0.5))
```
Note that in the right plot (when AH is calculated in the 0-48 h interval), the AH is closer to *p* than in the plot on the left.  
Due to this, it is important (at least inside each study), to always use the same time interval for *AH* calculation (we suggest to use *AH~24~*).






