---
title: "2. Milnesium anhydrobiosis"
author: "Matteo Vecchi"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
output: 
  workflowr::wflow_html:
    code_folding: hide 
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages, results='hide', message=F, warning=F}
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(ggstance)
library(RColorBrewer)
library(R2jags)
library(bayestestR)
library(xfun)    # Download file from html report
source("./code/AhI_functions.R")
```

We reanalyze the data from @Wojciechowska2021 about the effect of BHAM, a known AOX (mithochondrial alternative oxidase) inhibitor on the tardigrade *Milnesium inceptum*. Three different anhydrobiosis duration (3, 30 and 60 days) and 3 different BHAM  concentrations (control 0, 0.1 mM, 0.2 mM) were tested.  
The raw data was reformatted to be compatible with the anhydrobiosis index script, and can be downloaded here:
```{r echo=FALSE}
embed_file("./data/data_milnesium.txt")
```

```{r load data Milnesium}
data_Milnesium = read.table("./data/data_milnesium.txt", header=T, sep="\t")
data_Milnesium$total = rep(10, nrow(data_Milnesium)) # The total number of individuals analyzed for each replicate was 10
data_Milnesium$time_hours = data_Milnesium$time_min/60 # Convert time to hours
data_Milnesium$exp_group = paste(data_Milnesium$anhydrobiosis_time, data_Milnesium$treatment) # Merge together anhydrobiosis time and treatment
```

```{r run models, results='hide', message=F, warning=F, fig.dim = c(8, 8)}
# As it take times to run the model, we first check if we already run and saved it
ifelse("model_Milnesium.rds" %in% list.files("./output"),
       {model_Milnesium = readRDS("./output/model_Milnesium.rds")},
       {model_Milnesium = AhI_run_model(data_Milnesium, column_IDs = c(8,7,5,6))
        saveRDS(model_Milnesium, file="./output/model_Milnesium.rds")})

model_Milnesium_processed = AhI_analyze_model(model_Milnesium)

plot_AhI(model_Milnesium_processed)
```

```{r process model output}

```

```{r plot results, fig.width=8, fig.height=8}

```

## References


